The ``Dark Side'' of Phylogenetic Comparative Methods

### 2014-05-29 Manuscript/dark.tex
Finished introduction draft

Phylogenetic comparative methods (PCMs) were initially developed in the 1980s to deal with the statistical non-independence of species in comparative analyses. Since then PCMs have been extended to investigate evolutionary pattern and process, and include methods for investigating drivers of diversification, rates of trait evolution, and modes of evolution for morphological and ecological traits %list of citations or see Pennell & Harmon or O'Meara).

PCMs have become extremely popular over recent years; in 2013 alone \citet{harvey1991comparative}'s seminal book on the comparative method in evolutionary biology received 195 citations (Google Scholar search, 22nd May 2014), and at the 2014 Evolution meeting in Raleigh NC, we estimate that over X talks used PCMs. This increase in use is at least partially the result of the number of available R packages to run these analyses (for example ape, GEIGER, diversitree, caper - citations) and the community's increasing familiarity with R \citep{R-Core-Team:2014aa}. Unfortunately with this increase in popularity, comes an increase in misuse! % need to work on this paragraph

@@

@@

\section{Funding}
This work was supported by The European Commission CORDIS Seventh Framework Programme (FP7) Marie Curie CIG grant, proposal number: 321696 (NC)

\section{Acknowledgments}
Thanks to Star Wars for the hilarious jokes.

>>>>

Phylogenetic comparative methods (PCMs) were initially developed in the 1980s to deal with the statistical non-independence of species in comparative analyses (e.g. \citet{felsenstein1985phylogenies,grafen1989phylogenetic}. Since then PCMs have been extended to investigate evolutionary pattern and process, and include methods for investigating drivers of diversification, the tempo and mode of trait evolution, and models of speciation and extinction (see reviews in \citealp{o2012evolutionary} and \citealp{pennell2013integrative}).

PCMs have become extremely popular over recent years; in 2013 alone \citet{harvey1991comparative}'s book on the comparative method in evolutionary biology received 195 citations (Google Scholar, 22nd May 2014), and at the 2014 Evolution meeting in Raleigh NC, we estimate that over XX %need to count these!
talks used PCMs of one form or another. Unfortunately, PCMs make numerous assumptions and suffer from biases in the same way as any statistical method, but the more popular these methods become, the less awareness methods users seem to have of this ``dark side''. Thus assumptions and biases are often inadequately assessed in empirical studies, leading to poor model fits and mis- or over-interpreted results.

%Here we explore this problem in more detail.

%Such issues are the responsibility of end users but also of methods developers: the tools and approaches used to fit models are often far more user-friendly and better documented than the methods used to to assess whether that model fit is reasonable. To address these issues, we propose this symposium to discuss issues in both classical and recent PCMs, along with new research on detecting for these issues and accounting for them.  We hope that this will both increase awareness of these problems and encourage further research and careful thought in the area, along with better dialogue between method developers and method users.

%In this paper we discuss why we think this has happened, and suggest ways to bridge the knowledge gap between methods users and methods developers.

%This increase in use is at least partially the result of the number of available R packages to run these analyses (for example ape, GEIGER, diversitree, caper - citations) and the community's increasing familiarity with R \citep{R-Core-Team:2014aa}. Unfortunately with this increase in popularity, comes an increase in misuse! % need to work on this paragraph

@@

%Why are PCMs increasingly being misused? All evidence points to a knowledge-gap between methods users and methods developers.

\subsection{}

To apply a method correctly, methods users must first be aware of the assumptions and biases of the method in question.

- and how to detect whether their model adequately meets these assumptions - as well as details of the number of taxa required for sufficient power etc. If failure to account for the assumptions of the method is the result of a knowledge gap then there are several possible explanations:

\begin{enumerate}
\item The assumptions are not mentioned in the literature.
\item The assumptions are mentioned in the literature, but the literature is too dense and technical, or pertinent details are hidden within large sections of text.
\item The assumptions are mentioned in the literature but methods users have gone straight to the implementation of the method and the manual does not mention assumptions

Above we have assumed a knowledge gap, we are not considering cases where The method user is aware of the assumptions and how to test for them, but chooses not to because it means they cannot publish.

@@

\section{Funding}
This work was supported by The European Commission CORDIS Seventh Framework Program (FP7) Marie Curie CIG grant, proposal number: 321696 (NC)

\section{Acknowledgments}
Thanks to Star Wars for the ``hilarious'' jokes.
